Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-20 mr-review.Rmd
minor tweaks

## Statistical models and methods for MR with multiple instruments and one continuous outcome

@@

In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.

@@

@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].

MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).

@@

1. Relatively small proportion of SNPs have nonz
ero effects on the exposure  
1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.

@@

Researchers have also drawn on other generalized linear models for analysis in MR 
studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
power, with simulations of count random variables and binary random variables. With growing interest
in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].

@@

Modern GWAS for biobank-scale data offer the possibility
of discovering even more causal relationships among complex traits. The large
sample sizes in biobank studies, often greater than 100,000 
subjects, offer the possibility of uncovering associations of modest strength 
that might be undetectable in smaller studies. For example, the UK 
Biobank Study involves more than 500,000 subjects. Each UKB subject's 
data consists of millions of SNP genotypes and measurements for 
thousands of traits [@bycroft_uk_2018].

Other biobanks?

EMR & GWAS?

>>>>

## Statistical models and methods for MR with multiple instruments

@@

In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.

@@

@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID. MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).

@@

1. Relatively small proportion of SNPs have nonzero effects on the exposure  
1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.

@@

Researchers have also leveraged other generalized linear models for analysis in MR 
studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
power, with simulations of count random variables and binary random variables. With growing interest
in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].

@@

Modern GWAS for biobank-scale data offer the possibility
of discovering even more causal relationships among complex traits. The large
sample sizes in biobank studies, often greater than 100,000 
subjects, offer the possibility of uncovering associations of modest strength 
that might be undetectable in smaller studies. For example, the UK 
Biobank Study involves more than 500,000 subjects. Each UK Biobank subject's 
data consists of millions of SNP genotypes and measurements for 
thousands of traits [@bycroft_uk_2018].
